Concurrent Programming with CXL

### 2020-05-24 book/paper.tex
Playing with producer/consumer and semaphores
Updating some cxl output

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
The \texttt{process(self)} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

The main process waits for the two other processes to finish by polling
both \texttt{done} flags.
Finally, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.

@@

The following is a possible interleaving of processes~0 and~1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

Another type of CXL value is the \emph{atom}.  You have already
seen them: the names of constants, variables, and methods are
all atoms.  An atom is denoted using a period followed by the
name.  For example, the atom for the \texttt{process} method
is denoted \texttt{.process}.

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def process(self):
   0 Jump 12
   1 Frame ('process', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 13)] 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
==== Safety violation ====
__main__/() [0,12-31]           32 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process /0  [1-5]                6 dict{ .count:0, .done:dict{ 0:False, 1:False } }
process /1  [1-10]              11 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
process /0  [6-10]              11 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
__main__/() [32-34,37-42,44-48] 49 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL value.
\end{itemize}

A name tag consists of the name of the main method of the process,
along with the tag specified in the \texttt{spawn} statement.
In Figure~\ref{fig:inc}, the created processes have name tags
\texttt{process/0} and \texttt{process/1}.  The main process that
starts the execution has name tag \texttt{\_\_main\_\_/()}.

@@

>>>>

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

The following is a possible interleaving of incrementers~0 and~1:
\begin{enumerate}
\item incrementer 0 loads the value of \texttt{count}, which is 0;
\item incrementer 1 loads the value of \texttt{count}, which is still 0;
\item incrementer 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 sets \texttt{upDone} to \texttt{True};
\item incrementer 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

Another type of CXL value is the \emph{atom}.  It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
   0 Jump 39
   1 Frame ('incrementer', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 17)] 12
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Name ('count', 'Up.cxl', 2, 13)
   4 Load 1
   5 Nary ('+', 'Up.cxl', 2, 19) 2
   6 Name ('count', 'Up.cxl', 2, 5)
   7 Store 1
Up.cxl:3     done[self] = True;
   8 Constant (True, 'Up.cxl', 3, 18)
   9 LoadVar ('self', 'Up.cxl', 3, 10)
   10 Name ('done', 'Up.cxl', 3, 5)
   11 Store 2
   12 Return
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
==== Safety violation ====
incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/1 [1-11]        12 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
incrementer/0 [7-11]        12 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
main/() [14-19,22-27,29-36] 37 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL value.
\end{itemize}

A name tag consists of the name of the main method of the process,
along with an optional tag specified in the \texttt{spawn} statement.
The default tag is \texttt{()}.
In Figure~\ref{fig:inc}, the created processes have name tags
\texttt{incrementer/0}, \texttt{incrementer/1}, and \texttt{main/()}.

@@

\chapter{Semaphores}

So far we have looked at how to protect a single resource.  Sometimes
we have multiple but a limited amount of resources that require
protection.  That is, if there are only $n$ resources, no more than
$n$ processes can access them at a time.
